Engineering a Runtime System for AQL

### 2017-03-02 paper/paper.tex
testing conflicts
Updating the repository.

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,english,american,draftfoot]{ulb_thesis}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{url}
\usepackage{slashed}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{makeidx}
\makeindex

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}[chapter]
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

@@

\makeatother

@@

\begin{document}
\selectlanguage{english}%

\title{Engineering a Runtime System for AQL}

\author{Andrea Morciano}

\degreeaward{Docteur en XXX\\
(Département XXX, Faculté XXX)}
\maketitle
\begin{dedication}
To my parents.\end{dedication}
\begin{acknowledgments}
Thank you, my advisor!
\end{acknowledgments}
\vspace*{4cm}

@@

@@

Information Extraction (IE) is the discipline that addresses the task
of extracting structured information from unstructured sources automatically.
Sources usually take the form of text documents. The most explored
aspect of IE is the extraction of \emph{named entities.} Another main
topic, that has become object of research only in recent years, is
the extraction of \emph{relationships} between entities. IE is a field
with contributions from many different communities: Machine Learning,
Information Retrieval, Database, Web, Document Analysis. The interest
in towards it is motivated by the evergrowing amount of data that
is generated by our society, that is at a point in which manual analysis
is infeasible. Thus IE promises to bring value to many application
domains, most notably the enterprise world and the Web. There are
two main approaches to IE: \emph{rule-based }and \emph{statistical}.

@@

Extracting structured forms from flat data strings (containing, e.g.,
addresses). It allows more effective decuplication of information,
among the other things.

@@

Articles, conferences sites, individual research sites and similar
are explored to obain formatted citations of publications, later stored
in publicly accessible databases. They are capable to forward references
and may provide aggregate statistics and scoring information. See
for instance Google Scholar (\url{https://scholar.google.it/}).

@@

@@

We measure the accuracy of an IE task with two quantities:
\begin{itemize}
\item \emph{precision}: the percentage of correctly extracted entities among
the extracted entities;
\item \emph{recall}: the percentage of entities extracted among all the
existing entities in the input source.
\end{itemize}
The main difficulties to achieving a good level of accuracy are:
\begin{itemize}
\item the availability of a great set of clues that might be very different
in nature (e. g. orghographic properties, part of speech, typical
words, etc.) and that might be difficult to combine;
\item the difficulty of finding out missed extractions;
\item the fact that with the advancement of research the extracted data
structures keep increasing in complexity (for instance it is becoming
more difficult to identify the boundaries of an entity in a text).
\end{itemize}
While we are able to reach a very good level of accuracy for entity
extraction ($90\%$), relationship extraction is still quite unreliable
($50-70\%$ accuracy), mainly due to its intrinsic complexity, superior
to that of the task of entity extraction.

@@

In its initial phase IE was interesing only to research communities.
With the enormous amount of data they need to process, companies are
now concerned with IE too, and \emph{scalability} has become a central
issue, while many IE systems don't address the problem of efficiently
carrying out extraction tasks with sufficient attention. Efficiency
is influenced most notably by the efficiency of the following tasks:
\begin{itemize}
\item filtering documents in order to actually examine only the ones that
have good chances to contain the desired information;
\item focusing on the parts of a document that contain relevant information;
\item processing steps, like database querying, that are typically very
expensive and that might have to be performed on the selected pieces
of input.
\end{itemize}
Recently many solutions have been proposed to target the scalability
issue; one of these is indeed SystemT.

@@

Named entities are elements of interest in a text. Example of entities
are person names, street addresses, institutions, countries, and so
on. In this section and in the next one, I assume that the output
of an extraction task is a series of labels inserted into an input
text document (that becomes \emph{annotated}), although there are
other possibilities.

@@

\begin{equation}
\textrm{"Contextual Pattern}\longrightarrow\textrm{Action"}
\end{equation}
\\
Contextual patterns are a way to identify entitities by exploiting
their properties or the context in which they are usually met. The
most common formalism used to express them is that of \emph{regular
expressions} over tokens of the input. Actions mark the entities that
have been identified, and might consist in storing them in a database
or adding delimiters directly in the text.\\
Most systems in this category present a cascaded structure: an input
document goes through a series of processing phases, and the output
of a phase is the input of the next one. A famous example is that
of cascading grammars: formal grammars are evaluated in sequence on
the input.

@@

For the examples I use an idealized syntax, as in REF.
\begin{example}
Consider the task of identifying all mentions ISO standards in a text.
A rule for this purpose could be:

\begin{multline}
\left(\left\{ \textrm{String}=\textrm{"ISO"}\right\} \left\{ \textrm{String}=\textrm{"/IEC"}\right\} \left\{ \textrm{?}\right\} \left\{ \textrm{String}=\textrm{"/ASTM"}\right\} \left\{ \textrm{?}\right\} \right.\\
\left.\left\{ \textrm{Orthography type}=\textrm{number}\right\} \left\{ \left\{ \textrm{String}=\textrm{":"}\right\} \left\{ \textrm{Orthography type}=\textrm{number}\right\} \right\} \left\{ \textrm{?}\right\} \right)\\
\longrightarrow\textrm{ISO Standards}
\end{multline}

\end{example}

a
\begin{example}
Multiple entities can be matched at once. Imagine we need to find
mentions of (simple) street addresses in a text consisting of a street
name and a street number. A rule that matches the name and number
separatedly could be:

\begin{multline}
\left(\left\{ \textrm{Orthography type}=\textrm{mixed case word}\right\} \left\{ \textrm{*}\right\} \right)\textrm{: Name}\left(\left\{ \textrm{String}=\textrm{","}\right\} \right)\\
\left(\left\{ \textrm{Orthography type}=\textrm{number}\right\} \right)\textrm{: Number}\\
\longrightarrow\textrm{Street Name}=:\textrm{Name, }\textrm{Street Number}=:\textrm{Number}
\end{multline}

\end{example}

@@

\subsubsection{Token-Level Models\label{sub:Token-Level-Models}}

@@

Clues are of the form 
\begin{equation}
f:\left(x,y,i\right)\longmapsto\mathbb{R}
\end{equation}
where $x$ is a sequence of tokens, $i$ is a position in $x$ and
$y$ is a candidate label for the token at $i$. We distinguish between
these types of features:
\begin{itemize}
\item \emph{word features};
\item \emph{orthographic features;}
\item \emph{dictionary lookup features.}
\end{itemize}

\paragraph{Models for Labeling Token}

@@

In these models the label for a segment depends on the properties
of its token and on the previous segment. We can describe a segment
$s_{j}$ as:
\begin{equation}
s_{j}=\left(y_{i},l_{j},u_{j}\right)\label{eq:segment}
\end{equation}
where $y_{i}$ is the proposed label for $s_{j}$ and $l_{j},u_{j}$
are the start and end positions of $s_{j}$. Therefore, a feature
is of the form:
\begin{equation}
f\left(y_{i},y_{i-1},x,l_{j},u_{j}\right)
\end{equation}
where $x$ is the input sequence of tokens, $y_{i-1}$ is the proposed
label for the previous segment and the other symbols are defined as
for Equation \ref{eq:segment}. Besides token-level features (see
subsection \ref{sub:Token-Level-Models}), we can exploit the following
feature types:
\begin{itemize}
\item \emph{Similarity to an entity in a database;}
\item \emph{Length of the segment.}
\end{itemize}
There are also global segmentation models, that try to find the best
segmentation of a token sequence by maximizing a target function.

@@

When exctracting relationship between entities, we might face three
types of specific tasks:
\begin{itemize}
\item given a pair of entities, find the relationship between them;
\item given an entity $e$ and a relationship $r$, find all the other entities
$e'$ such that $\left(e,e'\right)\in r$;
\item given a big and open-ended input and a relationship $r$ find all
pairs of entities $e',e''$ such that $\left(e',e''\right)\in r$.
\end{itemize}

\subsection{Predicting the Relationship between a Entity Pair }

For the first task, we can exploit the following resources:
\begin{itemize}
\item \emph{surface tokens}: tokens that are usually placed between entities.
They are strong clues;
\item \emph{part of speech tags }(the most important being verbs);
\item \emph{syntactic parse tree}: allows grouping words in phrase types,
e.g. noun phrases, propositional phrases, and so on;
\item \emph{dependency graph}: it is a less expensive structure to compute
than the parse tree and it links a word to those that depend on it.
\end{itemize}
The main methods available to carry out the task are:
\begin{itemize}
\item \emph{Feature-based methods}, that simply transform the clues mentioned
above for usage by conventional classifier models;
\item \emph{Kernel-based methods}, that use kernel functions to encode the
similarity between two graphs;
\item \emph{Rule-based methods}, creating rules over structures around pairs
of entities.
\end{itemize}
The second task is a special case of the third one so I don't treat
it directly.

@@

In this chapter, I give a description of \emph{SystemT} and its extraction
rule language \emph{AQL} (Annotation Query Language), and of a new
way of evaluating AQL queries, that uses a formalism derived from
\emph{Finite State Automata}. Producing and analysing a concrete runtime
system for AQL queries that employs this new method is the object
of my thesis. Since SystemT is a rule-based system, I now focus only
on this category.

@@

As I mentioned in the previous chapter, scalability has now become
a central issue to IE. Companies now rely on IE for many tasks; a
prominent example is \emph{Business Intelligence}. Unfortunately,
many systems developed in the past don't address this concern correctly.
Traditionally, most rule-based system rely on cascading grammars:
formal grammars are executed in sequence on the input, each grammar
representing a stage that takes as input the ouput of the previous
one. Rules in such grammars are matched using regular expressions:
if a part of the input text satisfies the regular expression associated
with a rule, that rule is activated (i.e. ``fires'') and the corresponding
action is executed. Evaluating a grammar on a document tends to be
costly, because simply evaluating a single rule might require scanning
the whole document. On large datasets, the running time becomes enormous.
What's more, these kind of systems are not able to fulfill the expressivity
requirements of the complex extraction tasks.\\
SystemT was developed at IBM to address these issues. This system
is based on a new approach to extraction rules: the \emph{algebraic
approach}. Here, data manipulations are viewed as operations in a
(relational) algebra. Extraction tasks are defined using \emph{annotators}
(see Section \ref{sec:Entity-Extraction}) whose rules are viewed
as \emph{queries} on input documents, that act as virtual databases.
Complex annotators are obtained by combining simple ones by means
of \emph{relational operators}. By doing so, all the optimization
techniques used in regular databases become available, but new techniques
are possible too, due to the characteristics of text documents. Another
advantage of SystemT is that it is capable to handle \emph{overlapping
rule matchings}, due to rulels being fired independently, which are
difficult to resolve in cascading grammars, where they tipically require
disambiguation policies that only partially solve the problem.\\
I now give an overview of SystemT's high-level structure. Then, I
talk about its data model, execution model and algebra.

@@

\subsection{Data Model\label{sub:Data-Model}}

SystemT uses an \emph{object-relational} data model for annotations
on a text document, that allows applying logical operators over it,
that can be composed. There is an important assumption to mention
before continuing: an extraction task over a set of documents is performed
one document at a time. This means that any relationships between
entities in different documents are disregarded. This assumption is
crucial to some optimization techniques that SystemT uses. While there
exist tasks where considering these relationships would be useful
(think of the Web), still a large number of relevant tasks can be
carried out this way. In the following, I define the basic data types
of SystemT's data model. \\
I consider a finite alphabet $\varSigma$ of symbols (characters).
$\sum^{*}$ denotes the set of all strings of finite length over $\sum^{*}$.
A document is modeled as one such string. 
\begin{definition}
A document is a string $s\in\sum^{*}$.
\end{definition}

\noindent The most basic data type is the \emph{span}.
\begin{definition}
Given a string $s=\sigma_{1}...\sigma_{n}$ where $\forall i$, $\sigma_{i}\in\Sigma$,
with length$\left|s\right|=n$ and whose characters are indexed in
the natural way, a span is a substring $\left[i,j\right\rangle $
, where $i,j$ are indices of $s$ satisfying $1\leq i\leq j\leq n+1$.
A span of $s$ is also denoted as $s_{\left[i,j\right\rangle }$.
\end{definition}

\noindent Spans can be aggregated in \emph{tuples}, which are finite
sequences of spans\emph{. }Let us denote the set of all possible spans
of a string $s$ by $\textrm{Spans}\left(s\right)$ and by $\textrm{SVars}$
an infinite set of span variables, which can be assigned spans. A
$\left(V,s\right)-\textrm{tuple}$ is defined as follows:
\begin{definition}
Given a set $V\subseteq\textrm{SVars}$, $V$ being finite, and a
string $s\in\sum^{*}$, a $\left(V,s\right)-\textrm{tuple}$ is a
mapping $\mu:V\longmapsto\textrm{Spans}\left(s\right)$. When $V$
is clear from the context, we might call $\mu$ simply a $s-\textrm{tuple}$.
\end{definition}

\noindent A set of tuples is called a \emph{relation. }Here, I focus
on \emph{span relations}, which I formally define as $\left(V,s\right)-\textrm{relations}$.
\begin{definition}
\label{span-relation}A $\left(V,s\right)-\textrm{relation}$ is a
set of $\left(V,s\right)-\textrm{tuples}$. As before, we can speak
of $s-\textrm{relation}$ when $V$ is clear.
\end{definition}

\noindent Each operator in the algebra takes one or more span relations
as input and outputs a single span relation.

\subsection{Execution Model\label{sub:Execution-Model}}

@@

begin
\begin{enumerate}
\item $\left\{ \textrm{Read localDB into main memory}\right\} $
\item $R\longleftarrow E\left(\textrm{localDB}\right)$
\item $\left\{ \textrm{Add \ensuremath{R} to localDB}\right\} $
\item $\left\{ \textrm{Write modified localDB to disk}\right\} $
\end{enumerate}
\caption{\label{alg:Annotating-all-local}Annotating all local databases in
a global database (taken from REF).}

@@

\noindent In this procedure, step 2 in the for loop is the most expensive.

\subsection{Algebra of Operators \label{sub:Algebra-of-Operators}}

@@

Relational operators are the usual operators of relational algebra
that appear in classical database query plans. Important examples
are:
\begin{itemize}
\item \emph{selection} ($\sigma$);
\item \emph{projection }($\pi$);
\item \emph{Cartesian product }($\times$);
\item \emph{Union} ($\cup$);
\item \emph{Intersection} ($\cap$).
\end{itemize}

@@

Loosely speaking, span extraction operators take a pattern and a document
as input and output the maximal set of spans that match that pattern.
There are two main span extraction operators:
\begin{lyxlist}{00.00.0000}
\item [{\emph{standard~regular~expression~matcher~($\varepsilon_{re}$)}:}] this
operator takes a regular expression $r$ as input and it identifies
all non-overlapping matchings of $r$ in the current document, from
left to right.
\item [{\emph{dictionary~matcher~($\varepsilon_{d}$)}:}] this operator
outputs all the spans that match some entry in a given dictionary.
\end{lyxlist}
Although the dictionary matcher might seem useless since there is
a regular expression matcher, it has some advantages over it, as the
fact that it can find overlapping matchings or that it enforces the
semantics of word boundaries.

>>>>

%% LyX 2.2.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[11pt,english,american,draftfoot]{ulb_thesis}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{url}
\usepackage{pdfpages}
\usepackage{slashed}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{makeidx}
\makeindex

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
%% A simple dot to overcome graphicx limitations
\newcommand{\lyxdot}{.}

\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}[chapter]
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{tikz, pgf, pgfplots}
\usetikzlibrary{automata,graphs,positioning}
\usetikzlibrary{arrows,backgrounds,plotmarks}
\pgfplotsset{width=7cm,compat=newest}
\pgfplotsset{plot coordinates/math parser=false}

\@ifundefined{showcaptionsetup}{}{%
 \PassOptionsToPackage{caption=false}{subfig}}
\usepackage{subfig}
\makeatother

@@

\begin{document}
\includepdf{PageDeGarde}
\begin{dedication}
To my parents.
\end{dedication}
\begin{acknowledgments}
Thank you, my advisor!
\end{acknowledgments}
\vspace*{4cm}

@@

In this chapter, I give a generic introduction to Information Extraction.
This Chapter is a re-elaboration of \cite{Sarawagi:2008:IE:1498844.1498845}.
Please refer to the original source for a more comprehensive and detailed
overview. The figures and examples appearing in this Chapter are mine.

@@

Information Extraction (IE) is the discipline that addresses the task
of extracting structured information from unstructured sources automatically.
Sources usually take the form of text documents. The most explored
aspect of IE is the extraction of \emph{named entities.} Another main
topic, that has become object of research only in recent years, is
the extraction of \emph{relationships} between entities. IE is a field
with contributions from many different communities: Machine Learning,
Information Retrieval, Database, Web, Document Analysis. The interest
towards it is motivated by the constantly increasing amount of data
that is generated by our society. Manual analysis has become infeasible.
IE promises to bring value to many application domains, most notably
the enterprise world and the Web.

There are two prominent approaches to IE: 
\begin{description}
\item [{\emph{rule-based}}] which define a set of rules that the output
has to respect. The rules can be manually coded or learnt from examples;
\item [{\emph{statistical}}] which seek to identify a decomposition of
unstructured text and to label its components.
\end{description}

@@

Extracting structured forms from flat data strings (containing, e.g.,
addresses). It allows more effective deduplication of information,
among the other things.

@@

Articles, conference sites, individual research sites and similar
are explored to obain formatted citations of publications, later stored
in publicly accessible databases. They are capable to forward references
and may provide aggregate statistics and scoring information. See,
for instance, Google Scholar (\url{https://scholar.google.com/}).

@@

The main issues that have to be dealt with when performing IE can
be divided in two categories: \emph{accuracy} and \emph{running time}.

@@

We measure the accuracy of an IE task with two quantities:
\begin{itemize}
\item \emph{precision}: the percentage of correctly extracted entities among
the extracted entities;
\item \emph{recall}: the percentage of entities extracted among all the
existing entities in the input source.
\end{itemize}
The main difficulties to achieving a good level of accuracy are:
\begin{itemize}
\item the availability of a great set of clues that might be very different
in nature (e. g. orghographic properties, part of speech, typical
words, etc.) and that might be difficult to combine;
\item the difficulty of finding out missed extractions;
\item the fact that with the advancement of research the extracted data
structures keep increasing in complexity (for instance it is becoming
more difficult to identify the boundaries of an entity in a text).
\end{itemize}
While it is possible to reach a very good level of accuracy for entity
extraction ($90\%$), relationship extraction is still quite unreliable
($50-70\%$ accuracy), mainly due to its intrinsic complexity.

@@

In its initial phase IE was interesting only to research communities.
Nowadays companies, because of the enormous amount of data they need
to process on a regular basis, are concerned with IE too, and \emph{scalability}
has become a central issue, while many IE systems don't address the
problem of efficiently carrying out extraction tasks with sufficient
attention. Efficiency is influenced most notably by the efficiency
of the following tasks:
\begin{itemize}
\item filtering documents in order to actually examine only the ones that
have good chances to contain the desired information;
\item focusing on the parts of a document that contain relevant information;
\item processing steps, like database querying, that are typically very
expensive and that might have to be performed on the selected pieces
of input.
\end{itemize}
Recently many solutions have been proposed to target the scalability
issue; one of these is SystemT.

@@

Named entities are elements of interest in a text. Example of entities
are person names, street addresses, institutions, countries, and so
on. In this section and in the next one, I assume that the output
of an extraction task is a series of labels inserted into an input
text document (that becomes \emph{annotated}), although there are
other possible output formats.

@@

\begin{equation}
\textrm{"Contextual Pattern}\longrightarrow\textrm{Action"}
\end{equation}
Contextual patterns are a way to identify entitities by exploiting
their properties or the context in which they are usually met. The
most common formalism used to express them is that of \emph{regular
expressions} over tokens of the input. Actions mark the entities that
have been identified, and might consist in storing them in a database
or adding delimiters directly in the text.

Most systems in this category present a \emph{cascaded structure}:
an input document goes through a series of processing phases, and
the output of a phase is the input of the next one. A famous example
is that of \emph{cascading grammars}: formal grammars are evaluated
in sequence on the input.

@@

For the examples I use an idealized syntax, as in \cite{Sarawagi:2008:IE:1498844.1498845}.
\begin{example}
Consider the task of identifying all mentions ISO standards in a text.
A rule for this purpose could be:

\begin{multline}
\left(\left\{ \textrm{String}=\textrm{"ISO"}\right\} \left\{ \textrm{String}=\textrm{"/IEC"}\right\} \left\{ \textrm{?}\right\} \left\{ \textrm{String}=\textrm{"/ASTM"}\right\} \left\{ \textrm{?}\right\} \right.\\
\left.\left\{ \textrm{Orthography type}=\textrm{number}\right\} \left\{ \left\{ \textrm{String}=\textrm{":"}\right\} \left\{ \textrm{Orthography type}=\textrm{number}\right\} \right\} \left\{ \textrm{?}\right\} \right)\\
\longrightarrow\textrm{ISO Standards}
\end{multline}
\end{example}

\begin{example}
Multiple entities can be matched at once. Imagine we need to find
mentions of (simple) street addresses in a text consisting of a street
name and a street number. A rule that matches the name and number
separatedly could be:

\begin{multline}
\left(\left\{ \textrm{Orthography type}=\textrm{mixed case word}\right\} \left\{ \textrm{*}\right\} \right)\textrm{: Name}\left(\left\{ \textrm{String}=\textrm{","}\right\} \right)\\
\left(\left\{ \textrm{Orthography type}=\textrm{number}\right\} \right)\textrm{: Number}\\
\longrightarrow\textrm{Street Name}=:\textrm{Name, }\textrm{Street Number}=:\textrm{Number}
\end{multline}
\end{example}

@@

\subsubsection{Token-Level Models\label{subsec:Token-Level-Models}}

@@

The clues that these models exploit are of the form 
\begin{equation}
f:\left(\mathbf{x},y,i\right)\longmapsto\mathbb{R}
\end{equation}
where $\mathbf{x}$ is a sequence of tokens, $i$ is a position in
$\mathbf{x}$ and $y$ is a candidate label for the token at $i$.
We distinguish between these types of features:
\begin{itemize}
\item \emph{word features};
\item \emph{orthographic features;}
\item \emph{dictionary lookup features.}
\end{itemize}
\begin{figure}
Yesterday I watched a movie called ``The Matrix''

\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|}
\hline 
i & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10\tabularnewline
\hline 
\hline 
x & Yesterday & I & watched & a & movie & called & `` & The & Matrix & ``\tabularnewline
\hline 
y & $y_{1}$ & $y_{2}$ & $y_{3}$ & $y_{4}$ & $y_{5}$ & $y_{6}$ & $y_{7}$ & $y_{8}$ & $y_{9}$ & $y_{10}$\tabularnewline
\hline 
\end{tabular}

\caption{\label{fig:tokenization}Decomposition of a sentence into a sequence
of tokens.}
\end{figure}

\begin{example}
Consider the sentence shown in Figure \ref{fig:tokenization} and
its corresponding decomposition. An example of word feature at position
5 is

\[
f_{1}\left(y,\mathbf{x},i\right)=\left\llbracket x_{i}\textrm{ equals "Matrix"}\right\rrbracket \cdot\left\llbracket y=\textrm{Movie}\right\rrbracket 
\]
where $\left\llbracket P\right\rrbracket =1$ if predicate $P$ is
true, and 0 otherwise. An orthographic feature might be

\[
f_{2}\left(y,\mathbf{x},i\right)=\left\llbracket x_{i}x_{i+1}\textrm{ matches INITIAL\_QUOTE CapsWord}\right\rrbracket \cdot\left\llbracket y=\textrm{Movie}\right\rrbracket .
\]

Finally, an example of dictionary lookup feature is

\[
f_{3}\left(y,\mathbf{x},i\right)=\left\llbracket x_{i}\textrm{ in Movie\_dictionary}\right\rrbracket \cdot\left\llbracket y=\textrm{Movie}\right\rrbracket .
\]
\end{example}

\paragraph{Models for Labeling Tokens}

@@

In these models the label for a segment depends on the properties
of its tokens and on the previous segment. We can describe a segment
$s_{j}$ as:
\begin{equation}
s_{j}=\left(y_{i},l_{j},u_{j}\right)\label{eq:segment}
\end{equation}
where $y_{i}$ is the proposed label for $s_{j}$ and $l_{j},u_{j}$
are the start and end positions of $s_{j}$. Therefore, a feature
is of the form:
\begin{equation}
f\left(y_{i},y_{i-1},\mathbf{x},l_{j},u_{j}\right)
\end{equation}
where $\mathbf{x}$ is the input sequence of tokens, $y_{i-1}$ is
the proposed label for the previous segment and the other symbols
are defined as for Equation \ref{eq:segment}. Besides token-level
features (see Subsection \ref{subsec:Token-Level-Models}), we can
exploit the following feature types:
\begin{itemize}
\item \emph{Similarity to an entity in a database;}
\item \emph{Length of the segment.}
\end{itemize}
\begin{figure}
``The Matrix'' by The Wachowski Brothers

\begin{tabular}{|c|c|c|c|}
\hline 
i & %
\begin{tabular}{cccc}
1 & 2 & 3 & 4\tabularnewline
\end{tabular} & 5 & %
\begin{tabular}{ccc}
6 & 7 & 8\tabularnewline
\end{tabular}\tabularnewline
\hline 
\hline 
x & ``The Matrix'' & by & The Wachowski Brothers\tabularnewline
\hline 
$\left(l_{j},u_{j},y_{i}\right)$ & $1,4,M$ & $5,5,O$ & $6,8,D$\tabularnewline
\hline 
\end{tabular}

\caption{\label{fig:segmentation2}Segmentation of a sentence.}
\end{figure}

\begin{example}
Consider the sentence in Figure \ref{fig:segmentation2} and its corresponding
segmentation, which identifies a movie and a director. A basic feature
similarity feature for the director could be:

\[
f\left(y_{i},y_{i-1},\mathbf{x},6,8\right)=\left\llbracket x_{6}x_{7}x_{8}\textrm{ appears in a list of movie directors}\right\rrbracket \cdot\left\llbracket y_{i}=\textrm{Director}\right\rrbracket .
\]

A more realistic feature would make use of some similarity function,
rather than requiring an extact match. An example of length feature
is

\[
f\left(y_{i},y_{i-1},\mathbf{x},l,u\right)=\left\llbracket u-l=3\right\rrbracket \cdot\left\llbracket y_{i}=\textrm{Director}\right\rrbracket .
\]

\medskip{}
\end{example}

There exist also global segmentation models, that try to find the
best segmentation of a token sequence by maximizing a target function.

@@

When exctracting relationships between entities, we might face three
types of specific tasks:
\begin{enumerate}
\item given a pair of entities, find the relationship between them;
\item given an entity $e$ and a relationship $r$, find all the other entities
$e'$ such that $\left(e,e'\right)\in r$;
\item given a big and open-ended input and a relationship $r$ find all
pairs of entities $e',e''$ such that $\left(e',e''\right)\in r$.
\end{enumerate}

\subsection{Predicting the Relationship between an a Pair of Entities}

For the first task, we can exploit the following resources:
\begin{itemize}
\item \emph{surface tokens}: tokens that are usually placed between entities.
They are strong clues;
\item \emph{part of speech tags }(the most important being verbs);
\item \emph{syntactic parse tree}: allows grouping words in phrase types,
e.g. noun phrases, propositional phrases, and so on;
\item \emph{dependency graph}: it is a less expensive structure to compute
than the parse tree and it links a word to those that depend on it.
\end{itemize}
The main methods available to carry out the task are:
\begin{itemize}
\item \emph{Feature-based methods}, that simply transform the clues mentioned
above for usage by conventional classifier models;
\item \emph{Kernel-based methods}, that use kernel functions to encode the
similarity between two graphs;
\item \emph{Rule-based methods}, creating rules over structures around pairs
of entities.
\end{itemize}
The second task is a special case of the third, and won't be treated
here.

@@

In this chapter, I give a description of \emph{SystemT} and its extraction
rule language \emph{AQL} (Annotation Query Language), and of a new
way of evaluating AQL queries, that uses a formalism derived from
\emph{Finite State Automata}. This chapter is a re-elaboration from
\cite{4497502,Krishnamurthy:2009:SSD:1519103.1519105,Fagin:2015:DSF:2772377.2699442}.
In particular, the mathematical statements from \ref{first-stat}
to \ref{.last-stat} are taken or adapted from \cite{Fagin:2015:DSF:2772377.2699442},
except for Remarks \ref{first-rem} to \ref{last-rem}, that come
from \cite{4497502}. Please refer to the original sources for more
details on the topic. The figures and examples appearing in this chapter
are mine, except when I explicitly state otherwise.

@@

As I mentioned in the previous chapter, scalability is now a main
concern in IE. Companies rely on IE for many tasks; a prominent example
is \emph{Business Intelligence}. Unfortunately, many systems developed
in the past don't address this issue correctly. Traditionally, most
rule-based system rely on cascading grammars: formal grammars are
executed in sequence on the input, each grammar representing a stage
that takes as input the ouput of the previous one. Rules in such grammars
are matched using regular expressions: if a part of the input text
satisfies the regular expression associated with a rule, that rule
is activated (i.e. ``fires'') and the corresponding action is executed.
Evaluating a grammar on a document tends to be costly, because simply
evaluating a single rule might require scanning the whole document.
On large datasets, the running time becomes enormous. What's more,
these kind of systems are not able to fulfill the expressivity requirements
of the complex extraction tasks.

SystemT was developed at IBM to address these issues. This system
is based on a new approach to extraction rules: the \emph{algebraic
approach}. Here, data manipulations are viewed as operations in a
(relational) algebra. Extraction tasks are defined using \emph{annotators}
(see Section \ref{sec:Entity-Extraction}) whose rules are viewed
as \emph{queries} on input documents, that act as virtual databases.
Complex annotators are obtained by combining simple ones by means
of \emph{relational operators}. By doing so, all the optimization
techniques used in regular databases become available, but new techniques
are possible too, due to the characteristics of text documents. Another
advantage of SystemT is that it is capable to handle \emph{overlapping
rule matchings}, due to rulels being fired independently, which are
difficult to resolve in cascading grammars, where they tipically require
disambiguation policies that only partially solve the problem.

I now give an overview of SystemT's high-level structure. Then, I
talk about its data model, execution model and algebra. For more information
on SystemT and AQL, please refer to \cite{4497502,Krishnamurthy:2009:SSD:1519103.1519105}.

@@

\subsection{Data Model\label{subsec:Data-Model}}

SystemT uses an \emph{object-relational} data model for annotations
on a text document, that allows applying logical operators over it.
Operators can be composed. There is an important assumption to mention
before continuing: an extraction task over a set of documents is performed
on one document at a time. This means that any relationships between
entities in different documents are disregarded. This assumption is
crucial to some optimization techniques that SystemT uses. While there
exist tasks where considering these relationships would be useful
(think of the Web), still a large number of relevant tasks can be
carried out this way. In the following, I report the definitions of
the basic data types in the data model of SystemT.

I consider a finite alphabet $\Sigma$ of symbols (characters). $\Sigma^{*}$
denotes the set of all strings of finite length over $\Sigma$. A
document is modeled as one such string. 
\begin{definition}
\label{first-stat}A document is a string $\mathbf{s}\in\Sigma^{*}$.
\end{definition}

\noindent The most basic data type is the \emph{span}.
\begin{definition}
Given a string $\mathbf{s}=\sigma_{1}...\sigma_{n}$ where $\forall i$,
$\sigma_{i}\in\Sigma$, with length$\left|\mathbf{s}\right|=n$ and
whose characters are indexed in the natural way, a span is a substring
$\left[i,j\right\rangle $, where $i$, $j$ are indices of $s$ satisfying
$1\leq i\leq j\leq n+1$. A span of $\mathbf{s}$ is also denoted
as $\mathbf{s}_{\left[i,j\right\rangle }$.
\end{definition}

\noindent Spans can be aggregated in \emph{tuples}, which are finite
sequences of spans\emph{. }Let us denote the set of all possible spans
of a string $\mathbf{s}$ by $\textrm{Spans}\left(\mathbf{s}\right)$
and by $\textrm{SVars}$ an infinite set of span variables, which
can be assigned spans. A $\left(V,\mathbf{s}\right)-\textrm{tuple}$
is defined as follows:
\begin{definition}
Given a set $V\subseteq\textrm{SVars}$, $V$ being finite, and a
string $\mathbf{s}\in\sum^{*}$, a $\left(V,\mathbf{s}\right)-\textrm{tuple}$
is a mapping $\mu:V\longmapsto\textrm{Spans}\left(\mathbf{s}\right)$.
When $V$ is clear from the context, we might call $\mu$ simply a
$\mathbf{s}-\textrm{tuple}$.
\end{definition}

\noindent A set of tuples is called a \emph{relation. }Here, we focus
on \emph{span relations}, which are formally defined as $\left(V,\mathbf{s}\right)-\textrm{relations}$.
\begin{definition}
\label{span-relation}A $\left(V,\mathbf{s}\right)-\textrm{relation}$
is a set of $\left(V,\mathbf{s}\right)-\textrm{tuples}$. As before,
we can speak of $\mathbf{s}-\textrm{relation}$ when $V$ is clear.
\end{definition}

\noindent Each operator in the algebra of SystemT takes one or more
span relations as input and outputs a single span relation.

\subsection{Execution Model\label{subsec:Execution-Model}}

@@

begin
\begin{enumerate}
\item $\left\{ \textrm{Read localDB into main memory}\right\} $
\item $R\longleftarrow E\left(\textrm{localDB}\right)$
\item $\left\{ \textrm{Add \ensuremath{R} to localDB}\right\} $
\item $\left\{ \textrm{Write modified localDB to disk}\right\} $
\end{enumerate}
\caption{\label{alg:Annotating-all-local}Annotating all local databases in
a global database (taken from \cite{4497502}).}

@@

\subsection{Algebra of Operators \label{subsec:Algebra-of-Operators}}

@@

Relational operators are the usual operators of relational algebra
that appear in classical database query plans. Important examples
are:
\begin{itemize}
\item \emph{selection} ($\sigma$);
\item \emph{projection }($\pi$);
\item \emph{Cartesian product }($\times$);
\item Join ($\bowtie$);
\item \emph{Union} ($\cup$);
\item \emph{Intersection} ($\cap$).
\end{itemize}

@@

Loosely speaking, span extraction operators take a pattern and a document
as input and output a maximal set of spans that match that pattern.
There are two main span extraction operators:
\begin{lyxlist}{00.00.0000}
\item [{\emph{standard~regular~expression~matcher~($\varepsilon_{re}$)}:}] this
operator takes a regular expression $r$ as input and it identifies
all non-overlapping matchings of $r$ in the current document, from
left to right.
\item [{\emph{dictionary~matcher~($\varepsilon_{d}$)}:}] this operator
outputs all the spans that match some entry in a given dictionary.
\end{lyxlist}
Although the dictionary matcher might seem useless since there is
a regular expression matcher, it has some advantages over it, as the
fact that it can find overlapping matchings or that it enforces the
semantics of word boundaries.
